### 2010-08-19 Thesis/Chapter4.tex
- Update 19 August -
Chapter 4 finished (well, I'd like to believe it is)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@156 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

	\subsection{Meshless, position-based dynamics (supposedly good for handling topological changes but other issues)}

>>>>

	\subsection{Meshless methods (supposedly good for handling topological changes but other issues)}

If the finite element method is a robust method and widely used in engineering, the method has a few shortcomings \citep{Liu05}. First, the creation of a mesh is compulsory and producing a good quality mesh is both difficult and time consuming. Second, the stresses obtained in FEM are often discontinuous at the interfaces of the elements. Special techniques are required in a post-processing stage to recover accurate stresses. In addition, under large deformations, considerable loss in accuracy in FEM results can arise from the element distortions. The origin of these problems is the use of elements. Consequently, the idea of getting rid of the elements and the mesh has emerged and the concept of \emph{meshless} methods was born. Instead, meshless methods use a set of nodes scattered within the problem domain. They do not form a mesh, meaning it does not require any a priori information on the relationship between the nodes for the interpolation of the unknown variables. In theory, this should facilitate topological changes, a feature fairly desirable in surgery simulation. It is worth noting though that meshless methods are still in their developing stage and are not as mature as finite element formulations. We will now briefly review the research carried out in soft tissue modelling for medical simulation that makes use of meshless method. A complete review and details on meshless methods are beyond the scope of this thesis.

A meshless method was introduced for the first time for modelling soft tissue in medical simulation by \cite{De01}. They use the so-called method of finite spheres (MFS). Nodal points are only sprinkled locally around the surgical tool tip and not over the whole domain. The interpolation is performed by functions that are nonzero only on spheres surrounding the nodes. A point collocation technique is used to generate the discrete equations. The authors qualitatively compared the method of finite spheres with the finite element technique from a commercial package and the displacement profil was similar in both cases. They were able to achieved a computational rate of about $ 100\, $Hz when 34 spheres where used. Real-time haptic rates of about $ 1\, $kHz was then obtained using a force extrapolation technique.

\cite{Lim04} used the same technique of finite spheres introduced by \citeauthor{De01} to model soft tissue but they added progressive cutting, without the generation of new primitives. As the cut progresses, the nearest vertex to the intersection point of the tool is snapped to the tool path. Then, they observed that the prescribed boundary condition changes on only a very small portion as the tool interacts with the organ. Rather than solving the entire problem over and over again, they proposed instead to make incremental corrections, which results in an accelerated solution procedure.

Meshless methods were then introduced to computer graphics by \cite{Muller04}. In each step, they compute the spatial derivatives of the displacement field using a moving least squares (MLS) procedure. From these derivatives, they obtain strains, stresses and elastic forces at each simulated point. In addition, they proposed techniques for modeling and animating a point-sampled surface that dynamically adapts to deformations of the underlying volumetric model. They used a non-linear measure of strain with a linear constitutive model. They reported real-time simulation of models exibiting elastic, plastic, melting and flowing effects.

\cite{Horton06} tested meshless methods for surgical simulation. They proposed a total Lagrangian explicit dynamics algorithm using a non-linear material formulation. They applied the method on an example of brain shift and compared their results with a commercial finite element code. The shape functions are created from a cloud of unconnected nodes using the element free Galerkin method. They were interested in the displacement of the centreof mass for the brain and found the absolute displacement differences between the two algorithm are less than $ 0.85\,$mm which is the resolution of the MRIs in their study. The same team kept investigating and testing the technique. However, they used moving least squares procedure for creating the shape functions instead of the element free Galerkin method. They also realised experimental indentations \citep{Horton07} and provided more extensive comparisons with a finite element method \citep{Horton10}.

Another example of application is the used of the technique called moving particule semi-implicit method (MPS) by \cite{Chhatkuli09}. They applied this method to simulate the deformation of lung and a tumor inside the left lung during inspiration. The lung tissues were considered to be homogeneous, isotropic and viscoelastic. They compared their results with the experimental CT taken at the end of inspiration and foundout that the deformation predicted by numerical simulation matches reasonably well with the experimental results.

\TODO{add drawbacks of meshless methods}
